

---
title: "final_project_ryan"
output: html_document
date: "2024-03-07"
---

## Which type of road leads to the most injuries, what about fatalities? (RYAN)
## Which county has highest bike crash per CAP? (RYAN)
## Which county has highest fatalities per CAP?  (RYAN)

```{r}
library(dplyr)
library(tidyverse)
library(janitor)
library(ggplot2)
```

```{r}
bikes <- read_csv("datatesting/data/bicycle-fatality-and-injury-data-by-county.csv")
bikes <- clean_names(bikes) #names were upper case before
```

```{r}
county_data <- read_csv("datatesting/data/county_data.csv")
```
```{r}
county_data
```
```{r}
bikes
```

```{r}
bikes_new <- full_join(bikes, county_data)
```
```{r}
pop <- read_csv("datatesting/data/county_populations.csv")
```
```{r}
pop #found that county names do not match so asked chatgpt to fix county names
```

```{r}
updated_pop <- read_csv("datatesting/data/updated_county_populations.csv")
```
```{r}
updated_pop <- updated_pop %>%
  filter(county != "CALIFORNIA") #data has total pop which we dont need
```

```{r}
bikes_new <- full_join(bikes_new, updated_pop)
```
```{r}
bikes_new #population and year names are wrong
```

```{r}
bikes_new <- bikes_new %>%
  rename(year ="population")%>%
  rename(pop = "population.1")
```

```{r}
bikes_new
```





### finding what highway type has the highest numabr of injuries
```{r}
bikes_new %>% 
  group_by(highway_type) %>% 
  summarise(highest_injury_by_road = sum(injury)) %>% 
  arrange(desc=T) %>% 
  filter(highway_type != "TOTAL") %>% 
ggplot(aes(highway_type, highest_injury_by_road, fill = highway_type)) + geom_col(alpha = .7) + labs(title = "Injuries by Highway Type", x = "Highway Type", y= "Injuries", theme_minimal()) 
```
## plot for highest bike fatality per capita
```{r}
bikes_new %>% 
  mutate(death_per_cap = fatality/pop) %>% 
  arrange(desc(death_per_cap)) %>% 
  filter(highway_type == "TOTAL") %>% 
  head(death_per_cap, n=5) %>% 
 ggplot(aes(county,death_per_cap, fill = county)) +geom_col(alpha = .7) + labs(title = "Fatality per Capita by County", x= "County", y= "Fatality per Capita")
```
## highest crash rate per county
```{r}
bikes_new %>% 
  mutate(total_per_cap = total/pop) %>% 
  arrange(desc(total_per_cap)) %>% 
  filter(highway_type == "TOTAL") %>% 
  head(total_per_cap, n=5) %>% 
 ggplot(aes(county,total_per_cap, fill = county)) +geom_col(alpha = .7) + labs(title = "Crashes per Capita by County", x= "County", y= "Crashes per Capita")
```

### highest fatality per crash by county

```{r}
bikes_new %>% 
  mutate(total_per_cap = total/pop) %>% 
  mutate(death_per_cap = fatality/pop) %>% 
  mutate(death_per_crash_cap = death_per_cap/total_per_cap) %>% 
  arrange(desc(death_per_crash_cap)) %>% 
  filter(highway_type == "TOTAL") %>% 
  head(death_per_crash_cap, n=5) %>% 
  ggplot(aes(county,death_per_crash_cap, fill = county)) +geom_col(alpha = .7) + labs(title = "Fatalities per Crash by County", x= "County", y= "Fatality per Crash")
```

---
title: "County Coordinate Data"
author: Rohan
output: html_document
date: "2024-03-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load the Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(janitor)
library(shiny)
library(shinydashboard)
```

```{r}
bike_crashes <- read_csv("datatesting/data/bicycle-fatality-and-injury-data-by-county.csv") #bike injury data, part of vehicle crash data set
```

```{r}
bike_crashes$County <- as.factor(bike_crashes$County)
str(bike_crashes) #rohan
```

```{r}
levels(bike_crashes$County) #rohan
```

```{r}
# Created a data frame with the county names, latitudes, and longitudes using the help of ChatGPT
county_data <- data.frame(
  county = c("ALAMEDA", "ALPINE", "AMADOR", "BUTTE", "CALAVERAS", "COLUSA", "CONTRA COSTA", "DEL NORTE",
             "EL DORADO", "FRESNO", "GLENN", "HUMBOLDT", "IMPERIAL", "INYO", "KERN", "KINGS",
             "LAKE", "LASSEN", "LOS ANGELES", "MADERA", "MARIN", "MARIPOSA", "MENDOCINO", "MERCED",
             "MODOC", "MONO", "MONTEREY", "NAPA", "NEVADA", "ORANGE", "PLACER", "PLUMAS",
             "RIVERSIDE", "SACRAMENTO", "SAN BENITO", "SAN BERNARDINO", "SAN DIEGO", "SAN FRANCISCO",
             "SAN JOAQUIN", "SAN LUIS OBISPO", "SAN MATEO", "SANTA BARBARA", "SANTA CLARA", "SANTA CRUZ",
             "SHASTA", "SIERRA", "SISKIYOU", "SOLANO", "SONOMA", "STANISLAUS", "SUTTER", "TEHAMA",
             "TRINITY", "TULARE", "TUOLUMNE", "VENTURA", "YOLO", "YUBA"),
  latitude = c(37.6017, 38.5974, 38.3489, 39.6254, 38.196, 39.1789, 37.9191, 41.7435,
               38.7787, 36.9859, 39.5989, 40.745, 32.8397, 36.3093, 35.3433, 36.0741,
               39.1012, 40.6739, 34.0522, 36.9859, 38.0834, 37.4849, 39.5501, 37.2083,
               41.5885, 37.938, 36.2168, 38.5025, 39.1347, 33.7175, 38.9045, 40.0036,
               33.9534, 38.5816, 36.6115, 34.9592, 32.7157, 37.7749, 37.9577, 35.3102,
               37.563, 34.4208, 37.3541, 36.9741, 40.7909, 39.5774, 41.6639, 38.3105,
               38.5779, 37.5091, 39.0446, 40.1251, 40.6503, 36.1342, 37.8675, 34.3705, 38.7646, 39.2547),
  longitude = c(-121.7195, -119.8203, -120.7741, -121.537, -120.6805, -122.2342, -121.9283, -123.8974,
                -120.5231, -119.2321, -122.3935, -123.8695, -115.6121, -117.546, -118.7278, -119.8155,
                -122.7533, -120.5579, -118.2437, -120.5824, -122.7633, -119.9663, -123.4384, -120.6977,
                -120.7525, -118.8867, -121.2264, -122.2655, -121.171, -117.8311, -121.1448, -120.8393,
                -117.3962, -121.4944, -121.286, -116.4194, -117.1611, -122.4194, -121.2908, -120.4358,
                -122.3255, -119.6982, -121.9552, -122.0308, -121.8474, -120.5211, -122.545, -121.9018,
                -122.9888, -120.9876, -121.3153, -122.2345, -123.089, -118.8597, -120.2602, -119.1391, -121.9018, -121.3999)
)

# Print the data frame with county, latitude, and longitude
print(county_data) #rohan
```

# Which county has the highest fatality:crash ratio?
```{r}
# Calculate the fatality to crash ratio
bike_crashes$fatality_ratio <- bike_crashes$Fatality / bike_crashes$Total

# Find the county with the highest fatality:crash ratio
county_highest_ratio <- bike_crashes$County[which.max(bike_crashes$fatality_ratio)]

# Print the county with the highest fatality:crash ratio
print(county_highest_ratio) #rohan
```

```{r}
clean_names(county_data)
write.csv(county_data, "datatesting/data/county_data.csv", row.names = F) #kevin
```

#What is the mean fatality rate among different types of streets?
```{r}
# Calculate fatality rate for each type of street
fatality_rate <- bike_crashes %>%
  mutate(fatality_rate = Fatality / Total) %>%
  group_by(`Highway Type`) %>%
  summarise(mean_fatality_rate = mean(fatality_rate, na.rm = TRUE))

# Create a bar plot with enhanced aesthetics
ggplot(fatality_rate, aes(x = `Highway Type`, y = mean_fatality_rate)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Fatality Rate Among Types of Streets",
       x = "Highway Type",
       y = "Average Fatality Rate") +
  theme_minimal() +  # Minimalistic theme
  theme(
    plot.title = element_text(hjust = 0.5),  # Centering the title
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for better aesthetics
  ) #rohan
```