---
title: "final_project_ryan"
output: html_document
date: "2024-03-07"
---

## Which type of road leads to the most injuries, what about fatalities? (RYAN)
## Which county has highest bike crash per CAP? (RYAN)
## Which county has highest fatalities per CAP?  (RYAN)

```{r}
library(dplyr)
library(tidyverse)
library(janitor)
library(shiny)
#install.packages("ggmap")
library(ggmap)
library(RColorBrewer)
```

```{r}
bikes <- read_csv("datatesting/data/bicycle-fatality-and-injury-data-by-county.csv")
bikes <- clean_names(bikes) #names were upper case before
```

```{r}
county_data <- read_csv("datatesting/data/county_data.csv")
```
```{r}
county_data
```
```{r}
bikes
```

```{r}
bikes_new <- full_join(bikes, county_data)
```
```{r}
pop <- read_csv("datatesting/data/county_populations.csv")
```
```{r}
pop #found that county names do not match so asked chatgpt to fix county names
```

```{r}
updated_pop <- read_csv("datatesting/data/updated_county_populations.csv")
```
```{r}
updated_pop <- updated_pop %>%
  filter(county != "CALIFORNIA") #data has total pop which we dont need
```

```{r}
bikes_new <- full_join(bikes_new, updated_pop)
```
```{r}
bikes_new #population and year names are wrong
```

```{r}
bikes_new <- bikes_new %>%
  rename(year ="population")%>%
  rename(pop = "population.1")
```

```{r}
bikes_new
```





### finding what highway type has the highest numabr of injuries
```{r}
bikes_new %>% 
  group_by(highway_type) %>% 
  summarise(highest_injury_by_road = sum(injury)) %>% 
    arrange(desc=T) %>% 
  filter(highway_type != "TOTAL") %>% 
ggplot(aes(highway_type, highest_injury_by_road, fill = highway_type)) + geom_col(alpha = 1) + labs(title = "Injuries by Highway Type", x = "Highway Type", y= "Injuries") + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        plot.title = element_text(size = 14)) + scale_fill_brewer(palette = "Purples")
```
## plot for highest bike fatality per capita
```{r}
bikes_new %>% 
  mutate(death_per_cap = fatality/pop) %>% 
  arrange(desc(death_per_cap)) %>% 
  filter(highway_type == "TOTAL") %>% 
  head(death_per_cap, n=5) %>% 
 ggplot(aes(county,death_per_cap, fill = county)) +geom_col(alpha = .7) + labs(title = "Fatality per Capita by County", x= "County", y= "Fatality per Capita")+ scale_fill_brewer(palette = "YlGnBu")
```
## highest crash rate per county
```{r}
bikes_new %>% 
  mutate(total_per_cap = total/pop) %>% 
  arrange(desc(total_per_cap)) %>% 
  filter(highway_type == "TOTAL") %>% 
  head(total_per_cap, n=5) %>% 
 ggplot(aes(county,total_per_cap, fill = county)) +geom_col(alpha = 1) + labs(title = "Crashes per Capita by County", x= "County", y= "Crashes per Capita")+ scale_fill_brewer(palette = "BuGn")
```

### highest fatality per crash by county

```{r}
bikes_new %>% 
  mutate(total_per_cap = total/pop) %>% 
  mutate(death_per_cap = fatality/pop) %>% 
  mutate(death_per_crash_cap = death_per_cap/total_per_cap) %>% 
  arrange(desc(death_per_crash_cap)) %>% 
  filter(highway_type == "TOTAL") %>% 
  head(death_per_crash_cap, n=5) %>% 
  ggplot(aes(county,death_per_crash_cap, fill = county)) +geom_col(alpha = .7) + labs(title = "Fatalities per Crash by County", x= "County", y= "Fatality per Crash") + coord_flip() + scale_fill_brewer(palette = "Blues")
```
```{r}
bikes_new %>% 
  filter(highway_type == "TOTAL") %>% 
  filter(pop > 10000) %>% 
  ggplot(aes(pop, total)) + geom_point(color = "purple", alpha = .6) + scale_x_log10() + geom_smooth(method = "lm", color = "lavender") + labs(title = "Crashes by Population", x = "Population", y= "Crashes")
```
```{r}
bikes_new %>% 
  filter(county== "YOLO") %>% 
  filter(total != 0) %>% 
mutate(total_per_cap = total/pop) %>% 
  mutate(death_per_cap = fatality/pop) %>% 
  mutate(death_per_crash_cap = death_per_cap/total_per_cap)
```

#Created a data frame with the county names, latitudes, and longitudes using the help of ChatGPT
```{r}
county_data <- data.frame(
  county = c("ALAMEDA", "ALPINE", "AMADOR", "BUTTE", "CALAVERAS", "COLUSA", "CONTRA COSTA", "DEL NORTE",
             "EL DORADO", "FRESNO", "GLENN", "HUMBOLDT", "IMPERIAL", "INYO", "KERN", "KINGS",
             "LAKE", "LASSEN", "LOS ANGELES", "MADERA", "MARIN", "MARIPOSA", "MENDOCINO", "MERCED",
             "MODOC", "MONO", "MONTEREY", "NAPA", "NEVADA", "ORANGE", "PLACER", "PLUMAS",
             "RIVERSIDE", "SACRAMENTO", "SAN BENITO", "SAN BERNARDINO", "SAN DIEGO", "SAN FRANCISCO",
             "SAN JOAQUIN", "SAN LUIS OBISPO", "SAN MATEO", "SANTA BARBARA", "SANTA CLARA", "SANTA CRUZ",
             "SHASTA", "SIERRA", "SISKIYOU", "SOLANO", "SONOMA", "STANISLAUS", "SUTTER", "TEHAMA",
             "TRINITY", "TULARE", "TUOLUMNE", "VENTURA", "YOLO", "YUBA"),
  latitude = c(37.6017, 38.5974, 38.3489, 39.6254, 38.196, 39.1789, 37.9191, 41.7435,
               38.7787, 36.9859, 39.5989, 40.745, 32.8397, 36.3093, 35.3433, 36.0741,
               39.1012, 40.6739, 34.0522, 36.9859, 38.0834, 37.4849, 39.5501, 37.2083,
               41.5885, 37.938, 36.2168, 38.5025, 39.1347, 33.7175, 38.9045, 40.0036,
               33.9534, 38.5816, 36.6115, 34.9592, 32.7157, 37.7749, 37.9577, 35.3102,
               37.563, 34.4208, 37.3541, 36.9741, 40.7909, 39.5774, 41.6639, 38.3105,
               38.5779, 37.5091, 39.0446, 40.1251, 40.6503, 36.1342, 37.8675, 34.3705, 38.7646, 39.2547),
  longitude = c(-121.7195, -119.8203, -120.7741, -121.537, -120.6805, -122.2342, -121.9283, -123.8974,
                -120.5231, -119.2321, -122.3935, -123.8695, -115.6121, -117.546, -118.7278, -119.8155,
                -122.7533, -120.5579, -118.2437, -120.5824, -122.7633, -119.9663, -123.4384, -120.6977,
                -120.7525, -118.8867, -121.2264, -122.2655, -121.171, -117.8311, -121.1448, -120.8393,
                -117.3962, -121.4944, -121.286, -116.4194, -117.1611, -122.4194, -121.2908, -120.4358,
                -122.3255, -119.6982, -121.9552, -122.0308, -121.8474, -120.5211, -122.545, -121.9018,
                -122.9888, -120.9876, -121.3153, -122.2345, -123.089, -118.8597, -120.2602, -119.1391, -121.9018, -121.3999)
)

#rohan
```

```{r}
clean_names(county_data)
write.csv(county_data, "datatesting/data/county_data.csv", row.names = F) #kevin
```

#Which counties have the highest fatality rates?
```{r}

bikes_new <- bikes_new %>%
  mutate(fatality_ratio = fatality / total)


bikes_filtered <- bikes_new %>%
  filter(!is.na(fatality_ratio))


top_counties <- bikes_filtered %>%
  top_n(n = 6, wt = fatality_ratio)  # Select top 6 counties; adjust "n" as needed


county_colors <- viridis::viridis(nrow(top_counties))


ggplot(top_counties, aes(x = reorder(county, fatality_ratio), y = fatality_ratio, fill = county)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = county_colors) + 
  labs(title = "Top 5 Counties with Highest Fatality to Crash Ratio",
       x = "County",
       y = "Fatality to Crash Ratio") +
  theme_minimal() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8))  

```

#What is the mean fatality rate among different types of streets?
```{r}
bikes_filtered <- bikes_new %>%
  filter(highway_type != "TOTAL")

fatality_rate <- bikes_filtered %>%
  mutate(fatality_rate = fatality / total * 100) %>%
  group_by(highway_type) %>%
  summarise(mean_fatality_rate = mean(fatality_rate, na.rm = TRUE))

color_palette <- RColorBrewer::brewer.pal(8, "Set2")

ggplot(fatality_rate, aes(x = reorder(highway_type, mean_fatality_rate), y = mean_fatality_rate, fill = highway_type)) +
  geom_bar(stat = "identity", width = 0.6) +  
  scale_fill_manual(values = color_palette, name = "Highway Type") +  
  labs(title = "Average Fatality Rate Among Types of Streets",
       x = NULL,  
       y = "Average Fatality Rate (%)") +
  theme_minimal() +  
  theme(
    plot.title = element_text(size = 20, hjust = 0.5, margin = margin(b = 20)), 
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1, vjust = 1),  
    axis.text.y = element_text(size = 12),  
    axis.title.y = element_text(size = 14, margin = margin(t = 0, r = 10)),  
    axis.line = element_blank(),  
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank(),  
    panel.background = element_blank(),  
    panel.border = element_rect(color = "black", fill = NA, size = 0.5) 
  ) +
  coord_flip()  
```

## Is there a correlation between population size and fatalities? (ROHAN) 
```{r}
options(scipen = 999)

bikes_new %>%
  mutate(fatality_rate = fatality / total * 100) %>%
  filter(fatality_rate != 0) %>%
  ggplot(aes(x = pop, y = fatality_rate)) +
  geom_point(color = "blue", size = 2) +  
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed", size = 0.85) + 
  labs(title = "Correlation between Population Size and Fatality Rate",
       x = "Population",
       y = "Fatality Rate (%)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20, hjust = 0.5, margin = margin(b = 20)),  
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),  
    legend.title = element_text(size = 14), 
    legend.text = element_text(size = 12), 
    panel.border = element_rect(color = "black", fill = NA, size = 0.5)  
  ) +
  scale_x_log10()

```

```{r}
library(dplyr)
library(tidyverse)
library(janitor)
library(ggmap)
library(ggplot2)
library(maps)
library(shiny)
```

```{r}
bikes <- read_csv("datatesting/data/bicycle-fatality-and-injury-data-by-county.csv")
bikes <- clean_names(bikes) #names were upper case before
```

```{r}
county_data <- read_csv("datatesting/data/county_data.csv")
```
```{r}
county_data
```
```{r}
bikes
```

```{r}
bikes_new <- full_join(bikes, county_data)
```
```{r}
pop <- read_csv("datatesting/data/county_populations.csv")
```
```{r}
pop #found that county names do not match so asked chatgpt to fix county names
```

```{r}
updated_pop <- read_csv("datatesting/data/updated_county_populations.csv")
```
```{r}
updated_pop <- updated_pop %>%
  filter(county != "CALIFORNIA") #data has total pop which we dont need
```

```{r}
bikes_new <- full_join(bikes_new, updated_pop)
```
```{r}
bikes_new #population and year names are wrong
```

```{r}
bikes_new <- bikes_new %>%
  rename(year ="population")%>%
  rename(pop = "population.1")
```

```{r}
bikes_new
```

```{r}
bikes_capita <- bikes_new %>%
  mutate(total_capita = total/pop, fatality_capita = fatality/pop, injury_capita = injury/pop) #capita counts
view(bikes_capita)
```


```{r} #
ui <- fluidPage(
    titlePanel("Bike Crash Data in California by County"),
    sidebarLayout(
        sidebarPanel(
            selectInput("sizeColumn", "Select Column for Dot Size", 
                        choices = c("total", "fatality", "injury", "pop", "total_capita", "fatality_capita", "injury_capita"),
                        selected = "total")
        ),
        mainPanel(
            # Adjust the width and height as needed
            plotOutput("mapPlot", width = "800px", height = "600px")
        )
    )
)

server <- function(input, output) {
  output$mapPlot <- renderPlot({
    # Define the base map; here we're using a simple world map. For more detailed maps, you would use get_map().
    base_map <- ggplot() + borders("state", colour = "gray", fill = "white") +
      theme_minimal() +
      coord_fixed(ratio = 1.3) +
      xlim(-124.48, -113.94) + ylim(32.53, 42.01) # These limits are approximate for California
    
    # Prepare the data for plotting
    # Adjust the factor for dot size to fit your data's scale
    data_for_plot <- bikes_capita %>%
      mutate(size = get(input$sizeColumn) * 100) # You might need to adjust this scaling factor
    
    # Add points and text to the map
    map <- base_map +
      geom_point(data = data_for_plot, aes(x = longitude, y = latitude, size = size)) +
      geom_text(data = data_for_plot, aes(x = longitude, y = latitude, label = county), vjust = -1, size = 1.75) +
      labs(size = input$sizeColumn) # Dynamic label for dot size based on selection
    
    # Print the map
    print(map)
  })
}

shinyApp(ui, server)
```

```{r} #


ui <- fluidPage(
    selectInput("y", "Select Y Variable", 
                choices = c("total", "fatality", "injury", "pop", "total_capita", "fatality_capita", "injury_capita"), 
                selected = "total"),
    plotOutput("plot", width = "100%", height = "600px") # Use percentage for responsive width
)

server <- function(input, output) {
  output$plot <- renderPlot({

    # Dynamically reorder the counties based on the selected variable in descending order
    data_ordered <- bikes_capita %>%
      mutate(county_reordered = reorder(county, get(input$y), FUN = mean, decreasing = TRUE)) # Use decreasing = TRUE for descending order
    
    # Use the reordered data to plot
    ggplot(data_ordered, aes(x = county_reordered, y = get(input$y))) + 
      geom_bar(stat = "identity", fill = "steelblue") + # Added fill color for bars
      coord_flip() +
      theme_light(base_size = 10) + # Set base theme
      theme(axis.text.y = element_text(size = 7)) + # Adjust y-axis text size appropriately
      labs(title = "County Bike Crash Data", y = input$y, x = "County") # Add plot and axis titles
  })
}
shinyApp(ui, server)
```